---
title: "KM plots based on TBXT mRNA levels"
author: "Helena"
date: "08/09/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Before starting the analysis

Load the data and store it in an object 'samples'. The data I am using was downloaded from the UCSC Xena project. It combines RNASeq data from TCGA (cancer tissues) and GTex (normal tissues) via a common bioinformatics pipeline - so you can directly compare values from both studies.

```{r}
samples <- 
  read.delim("~/Desktop/crc_prognosis/tcga_target_gtex_colorectal.tsv", 
             header=TRUE)
```

Load required packages

```{r, message=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(forcats)
```

## Do some preliminery analyses

Summarise numbers of samples of each type in your dataset. We are only interested in 'Primary tumor'(which comes from the TCGA dataset) and 'Normal Tissue' (GTex) for the following analyses. 

(I have decided not to include the 'Solid Tissue Normal' samples in the following analyses. These are from the TCGA and are normal samples adjacent to some of the tumor samples as opposed to from non-cancer patients and therefore not independent data points. I should do a boxplot of these vs their primary tumor counterparts though.)

```{r}
count(samples, X_sample_type)
```

This is a boxplot summarising TBXT mRNA levels in normal colon tissue (GTex) vs primary colorectal adenoma.

```{r}
normal <- filter(samples, X_sample_type == "Normal Tissue")
cancer <- filter(samples, X_sample_type == "Primary Tumor")
boxplot(normal$T, cancer$T,
        at = c(1,2),
        names = c("Normal", "Cancer"),
        xlab = "Tissue type",
        ylab = "T expression (log2(norm_count+1))")
```

Calculate percentage of Brachyury positive samples in the normal tissue and cancer tissue

```{r}
(count(normal, T >0))/(count(normal))*100
```
