---
title: "KM plots based on TBXT mRNA levels"
author: "Helena"
date: "08/09/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
samples <- 
  read.delim("~/Desktop/crc_prognosis/tcga_target_gtex_colorectal.tsv", 
             header=TRUE)
```

```{r, include=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(forcats)
```

## Boxplot shows distributuon of data

This is a boxplot summarising TBXT mRNA levels in normal colon tissue (GTex) vs primary colorectal adenoma (TCGA).

```{r}
normal <- filter(samples, X_sample_type == "Normal Tissue")
cancer <- filter(samples, X_sample_type == "Primary Tumor")
boxplot(normal$T, cancer$T,
        at = c(1,2),
        names = c("Normal", "Cancer"),
        xlab = "Tissue type",
        ylab = "T expression (log2(norm_count+1))")
```

As you can see from the boxplot although the majority of samples in both datasets are similarly distributed with no/ low TBXT mRNA, the cancer dataset has more outliers that exhibit higher TBXT expression.

## Survival analyses (Kaplan Meier plots)

These analyses define a high Brachyury group and a low Brachyury group, look at the relative survival of those groups then test for a significant different in survival using a logrank test. You get different results depending on how you define the groups - where you put the cut-off between low and high. Here I go through what the results are for lots of different definitions.

## Above the max normal value

The max value for TBXT expression in the normal tissues (top dot in the boxplot) is:

```{r}
max(normal$T)
```
If we define 'High Brachyury' to be > 4.411 the analysis looks like this.
```{r}
cancer <- cancer %>% 
  mutate(T_group = ifelse(T > 4.4110, "> normal range","normal range"))
cancer$T_group <- factor(cancer$T_group)
surv_object <- Surv(time = cancer$OS.time, event = cancer$OS)
fit1 <- survfit(surv_object ~ T_group, data = cancer)
ggsurvplot(fit1, data = cancer, pval = TRUE, xlab = "Time(days)")
```

There appears to be a difference between the groups but it narrowly misses significance in the log-rank test.
This table shows the number of people in each group and the number of 'events' (by which it means death).
```{r}
fit1
```

## Top quartile (25%)

When we define 'High Brachyury' as the top quartile the analysis is as follows.
Find top quartile:

```{r}
summary(cancer$T)
```
Make the Kaplan-Meier plot:
```{r}
cancer <- cancer %>% 
  mutate(T_group_quart = ifelse(T >= 1.1160, "High","Low"))
cancer$T_group_quart <- factor(cancer$T_group_quart)
surv_object <- Surv(time = cancer$OS.time, event = cancer$OS)
fit1 <- survfit(surv_object ~ T_group_quart, data = cancer)
ggsurvplot(fit1, data = cancer, pval = TRUE, xlab = "Time(days)")
fit1
```

There is no difference between the groups. I don't think this is particularly suprising because lots of these samples have TBXT values well within the range of the normal tissues. 

## Arbitary cut-offs

So, starting from the cut-off of 4.4110 what happens if we bump the cut-off upwards slightly? First to 4.5.
```{r}
cancer <- cancer %>% 
  mutate(T_group_arb = ifelse(T > 4.5, "High","Low"))
cancer$T_group_arb <- factor(cancer$T_group_arb)
surv_object <- Surv(time = cancer$OS.time, event = cancer$OS)
fit1 <- survfit(surv_object ~ T_group_arb, data = cancer)
ggsurvplot(fit1, data = cancer, pval = TRUE, xlab = "Time(days)")
fit1
```
This loses only 1 patient from the high group and the results are now significant. 

Using 4.6 as a cut-off we 'lose' one more patient from the 'High' group and it looks like this:
```{r}
cancer <- cancer %>% 
  mutate(T_group_arb = ifelse(T > 4.6, "High","Low"))
cancer$T_group_arb <- factor(cancer$T_group_arb)
surv_object <- Surv(time = cancer$OS.time, event = cancer$OS)
fit1 <- survfit(surv_object ~ T_group_arb, data = cancer)
ggsurvplot(fit1, data = cancer, pval = TRUE, xlab = "Time(days)")
fit1
```
There is a more significant difference between the groups.

If I do this for lots of different arbitary cutoffs (between high and low Brachyury) - the P value changes as follows:

|Cut-off |P value |n |Events |
|:-------|:-------|:-|:------|
|4.411   |0.58    |19|6      |
|4.5     |0.042   |18|6      |  
|4.6     |0.032   |17|6      |
|4.7     |0.12    |16|5      |
|4.8     |0.078   |15|5      |
|5.1     |0.056   |14|5      |
|5.3     |0.02    |12|5      |
|5.5     |0.0075  |11|5      |
|5.7     |0.0027  |10|5      |
|5.9     |0.026   |9 |4      |
|6.5     |0.16    |8 |3      |
|6.7     |0.39    |6 |2      |

The cutoff which gives the lowest P value is 5.7 and the plot looks like this. 

```{r}
cancer <- cancer %>% 
  mutate(T_group_arb = ifelse(T > 5.7, "High","Low"))
cancer$T_group_arb <- factor(cancer$T_group_arb)
surv_object <- Surv(time = cancer$OS.time, event = cancer$OS)
fit1 <- survfit(surv_object ~ T_group_arb, data = cancer)
ggsurvplot(fit1, data = cancer, pval = TRUE, xlab = "Time(days)")
fit1
```

## Conclusion
There does seem to be a difference in prognosis between patients with >normal and normal levels of TBXT mRNA. The observation of a difference seems to be robust at many different cut-offs, but the level of statistical significance varies depending on the cut-off. 


